### 2025-09-14 YEAR2/MAT292_LectureNotes/main.tex
push

\begin{example}
    Consider the ODE:
    $$
    x' = rx + x^3 - x^5
    $$
    We can find the fixed points by setting \( x' = 0 \), and we have
    $$
    x^\star = 0 \quad \text{or} \quad x^\star = \pm\sqrt{\frac{1 \pm \sqrt{1 + 4r}}{2}}
    $$
    Now, we see the bifurcation points are at \( r = -\frac{1}{4} \) and \( r = 0 \). We have:
    \begin{itemize}
        \item For \( r < -\frac{1}{4} \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
        \end{itemize}
        \item For \( r = -\frac{1}{4} \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
            \item \( x^* = \pm\frac{1}{2} \) are semi-stable (since \( f'(\pm\frac{1}{2}) = 0 \))
        \end{itemize}
        \item For \( -\frac{1}{4} < r < 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
            \item \( x^* = -\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(-\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
            \item \( x^* = \sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is unstable (since \( f'(\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) > 0 \))
        \end{itemize}
        \item For \( r = 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is semi-stable (since \( f'(0) = 0 \))
            \item \( x^* = -1 \) is stable (since \( f'(-1) < 0 \))
            \item \( x^* = 1 \) is stable (since \( f'(1) < 0 \))
        \end{itemize}
        \item For \( r > 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is unstable (since \( f'(0) = r > 0 \))
            \item \( x^* = -\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(-\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
            \item \( x^* = \sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
        \end{itemize}
    \end{itemize}
    The bifurcation diagram is as follows:
    \begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \begin{axis}[
            axis lines = middle,
            xlabel = {$r$},
            ylabel = {$x^\star$},
            ymin = -2, ymax = 2,
            xmin = -1, xmax = 1,
            samples=200,
            domain=-0.25:1,
            legend style={at={(1.05,1)}, anchor=north west}
        ]
        % Stable and unstable branches from formula
        \addplot[blue, thick, domain=-0.25:1] {sqrt((1 + sqrt(1+4*x))/2)};
        \addplot[red, dashed, domain=-0.25:1] {sqrt((1 - sqrt(1+4*x))/2)};
        \addplot[blue, thick, domain=-0.25:1] {-sqrt((1 + sqrt(1+4*x))/2)};
        \addplot[red, dashed, domain=-0.25:1] {-sqrt((1 - sqrt(1+4*x))/2)};

@@

        \legend{stable, unstable}
        \end{axis}
    \end{tikzpicture}
    \caption{Bifurcation diagram for $x' = rx + x^3 - x^5$. 
    Blue = stable equilibria, red dashed = unstable equilibria.}
\end{figure}

>>>>

\begin{example}
    Consider the ODE:
    $$
    x' = rx + x^3 - x^5
    $$
    We can find the fixed points by setting \( x' = 0 \), and we have
    $$
    x^\star = 0 \quad \text{or} \quad x^\star = \pm\sqrt{\frac{1 \pm \sqrt{1 + 4r}}{2}}
    $$
    Now, we see the bifurcation points are at \( r = -\frac{1}{4} \) and \( r = 0 \). We have:
    \begin{itemize}
        \item For \( r < -\frac{1}{4} \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
        \end{itemize}
        \item For \( r = -\frac{1}{4} \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
            \item \( x^* = \pm\frac{1}{2} \) are semi-stable (since \( f'(\pm\frac{1}{2}) = 0 \))
        \end{itemize}
        \item For \( -\frac{1}{4} < r < 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
            \item \( x^* = -\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(-\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
            \item \( x^* = \sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is unstable (since \( f'(\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) > 0 \))
        \end{itemize}
        \item For \( r = 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is semi-stable (since \( f'(0) = 0 \))
            \item \( x^* = -1 \) is stable (since \( f'(-1) < 0 \))
            \item \( x^* = 1 \) is stable (since \( f'(1) < 0 \))
        \end{itemize}
        \item For \( r > 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is unstable (since \( f'(0) = r > 0 \))
            \item \( x^* = -\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(-\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
            \item \( x^* = \sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
        \end{itemize}
    \end{itemize}
    The bifurcation diagram is as follows:
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \begin{axis}[
                axis lines = middle,
                xlabel = {$r$},
                ylabel = {$x^\star$},
                ymin = -2, ymax = 2,
                xmin = -1, xmax = 1,
                samples=200,
                domain=-0.25:1,
                legend style={at={(1.05,1)}, anchor=north west}
            ]
            % Stable and unstable branches from formula
            \addplot[blue, thick, domain=-0.25:1] {
                sqrt((1 + sqrt(1+4*x))/2)
            };
            \addplot[red, dashed, domain=-0.25:1] {
                sqrt((1 - sqrt(1+4*x))/2)
            };
            \addplot[blue, thick, domain=-0.25:1] {
                -sqrt((1 + sqrt(1+4*x))/2)
            };
            \addplot[red, dashed, domain=-0.25:1] {
                -sqrt((1 - sqrt(1+4*x))/2)
            };

@@

            \legend{stable, unstable}
            \end{axis}
        \end{tikzpicture}
        \caption{Bifurcation diagram for $x' = rx + x^3 - x^5$.}
    \end{figure}
\end{example}

\paragraph{Conditions for Bifurcations} Consider $x' = f(x, r)$. Suppose ($x^c, r^c$) is a candidate for a bifurcation point. Then, the following conditions must hold:
\begin{itemize}
    \item \textbf{Equilibrium Condition} $f(x^c, r^c) = 0$ (i.e., $x^c$ is a fixed point for parameter value $r^c$)
    \item \textbf{Non Hyperbolicity Condition} $f_x(x^c, r^c) = 0$ (i.e., the Jacobian at the fixed point has a zero eigenvalue)    
    \item \textbf{Transversality Condition} $f_r(x^c, r^c) \neq 0$ (i.e., the fixed point changes with respect to the parameter)
    \item \textbf{Types of Bifurcations} For different types of bifurcations, we have the following additional conditions:
    \begin{itemize}
        \item \textbf{Saddle-Node Bifurcation} $f_{xx}(x^c, r^c) \neq 0$        
        \item \textbf{Pitchfork Bifurcation} $f_{xx}(x^c, r^c) = 0$ and $f_{xxx}(x^c, r^c) \neq 0$
    \end{itemize}
\end{itemize}
\subsection{Existence and Uniqueness of Solutions}
\begin{example}
    Consider the following IVP:
    $$
    y' = y^{\frac{1}{3}}, \quad y(0) = 0
    $$
    Apart from the trivial solution \( y(t) = 0 \), we can also solve this ODE by separating variables:
    \begin{align*}
        \frac{dy}{y^{1/3}} &= dt \\
        \int y^{-1/3} \, dy &= \int 1 \, dt \\
        \frac{3}{2} y^{2/3} &= t + C \\
        \intertext{Using the initial condition \( y(0) = 0 \), we find \( C = 0 \). Thus, the solution is:}
        y(t) &= \left( \frac{2}{3}(t ) \right)^{3/2}
    \end{align*}
    Thus, we have two solutions to the IVP. Existence is guaranteed, but uniqueness is not.
\end{example}

\begin{example}
    Consider the following IVP:
    $$
    y' = y^2 , \quad y(0) = 1
    $$
    We can solve this ODE by separating variables:
    \begin{align*}
        \frac{dy}{y^2} &= dt \\
        \int y^{-2} \, dy &= \int 1 \, dt \\
        -y^{-1} &= t + C \\
        \intertext{Using the initial condition \( y(0) = 1 \), we find \( C = -1 \). Thus, the solution is:}
        y(t) &= \frac{1}{1 - t}
    \end{align*}
    The solution exists but not globally, as it blows up at \( t = 1 \). Due to the nature of the initial condition, the solution is defined only for \( t < 1 \) to guarantee continuousness.
\end{example}

\begin{theorem}[Existence and Uniqueness Theorem for Linear ODEs]
    Consider the IVP:
    $$
    y' + p(t)y = g(t), \quad y(t_0) = y_0
    $$
    where \( p(t) \) and \( g(t) \) are continuous on an open interval \( I = (\alpha, \beta) \) containing \( t_0 \). Then, there exists a unique solution \( y(t) \) defined on the entire interval \( I \).
\end{theorem}
\begin{proof}
    (\textbf{Existence}) We can use the integrating factor method to find a solution. The integrating factor is given by:
    $$\mu(t) = \exp\left( \int p(t) \, dt \right)
    $$
    Multiplying both sides of the ODE by the integrating factor, we have:
    $$\mu(t) y' + \mu(t) p(t) y = \mu(t) g(t)
    $$
    Integrating both sides with respect to \( t \):
    $$\mu(t)y = \int \mu(s) g(s) \, ds + C
    $$
    We can determine \( C \) using the initial condition \( y(t_0) = y_0 \):
    $$y_0 = \frac{1}{\mu(t_0)}\left( \int_{t_0}^{t_0} \mu(s) g(s) \, ds + y_0 \right)
    $$
    Thus, we have a well defined solution \( y(t) \) defined on the interval \( I \), as long as $p$ and $g$ are continuous on $I$.

    (\textbf{Uniqueness}) Suppose there are two solutions \( y_1(t) \) and \( y_2(t) \) to the IVP, then they must satisfy $y_1(t_0) = y_2(t_0) = y_0$. Define \( z(t) = y_1(t) - y_2(t) \), then \( z(t) \) satisfies:
    $$
    z' + p(t)z = g - g = 0, \quad z(t_0) = y_1(t_0) - y_2(t_0) = 0
    $$
    Apply the same integrating factor method, we have:
    $$\mu(t) z' + \mu(t) p(t) z = 0
    $$
    Integrating both sides with respect to \( t \):
    $$\mu(t)z = \int_{t_0}^{t} 0 \, ds + C = C
    $$
    Using the initial condition \( z(t_0) = 0 \), we find \( C = 0 \). Then, since $\mu$ is a exponential, $\mu > 0$, \( z(t) = 0 \) for all \( t \in I \), which implies \( y_1(t) = y_2(t) \) for all \( t \in I \). Therefore, the solution to the IVP is unique.
\end{proof}

\begin{theorem}[Cauchy-Lipschitz Theorem]
    Consider the IVP:
    $$
    y' = f(t, y), \quad y(t_0) = y_0
    $$
    where \( f(t, y) \) and \( f_y(t, y) \) are continuous on a rectangle \( R = [\alpha, \beta] \times [\gamma, \delta] \). Then, for some point \((t_0, y_0) \in \mathrm{int}(R)\), there exists an interval \( I = (t_0 - h, t_0 + h) \subseteq [\alpha, \beta] \) for some \( 0 < h \le \min(t_0 - \alpha, \beta - t_0) \), such that there exists a unique solution \( y(t) \) defined on the interval \( I \).

\end{theorem}

\begin{proof}
    The proof involves the method of successive approximations (Picard iterations). Which we will not cover here.
\end{proof}

\begin{example}
    To demonstrate that the Cauchy-Lipschitz theorem applies in the linear case, consider the IVP:
    $$
    y' + p(t)y = g(t), \quad y(t_0) = y_0
    $$
    where \( p(t) \) and \( g(t) \) are continuous. We also have \( f_y = -p(t)y + g(t) = -p(t) \). So both \( f \) and \( f_y \) are continuous, thus the Cauchy-Lipschitz theorem guarantees the existence and uniqueness of the solution on some interval around \( t_0 \).
\end{example}

\begin{example}
    Consider the IVP:
    $$
    y' = y^{1/3}, \quad y(0) = 0
    $$
    Here, \( f(t, y) = y^{1/3} \) and \( f_y(t, y) = \frac{1}{3}y^{-2/3} \). The function \( f(t, y) \) is continuous everywhere, but \( f_y(t, y) \) is not continuous at \( y = 0 \). Therefore, the Cauchy-Lipschitz theorem does not apply at the initial condition \( (0, 0) \), which explains why we have multiple solutions to this IVP.
\end{example}

### 2025-09-14 YEAR2/PHY293_LectureNotes/main.tex
push

\subsection{Energy and Quality Factor}
\begin{definition}[Energy of a Very Light Damping]
    Consider a very lightly damped harmonic oscillator, where \( \gamma \ll \omega_0 \). In this case, the angular frequency of oscillation \( \omega_r \) can be approximated as:
    $$
    \omega_r \approx \omega_0 \left(1 - \frac{\gamma^2}{8\omega_0^2}\right) \approx \omega_0
    $$
    So the motion of the lightly damped oscillator can be approximated as:
    $$
    x(t) \approx A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)
    $$
    Then, we can calculate the velocity of the oscillator:
    \begin{align*}
        \dot{x}(t) &= -\frac{\gamma}{2} A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi) - A_0 \omega_0 e^{-\frac{\gamma}{2}t} \sin(\omega_0 t + \phi) \\
        &= A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)
    \end{align*}
    The total mechanical energy \( E(t) \) of the lightly damped oscillator is given by:
    \begin{align*}
        E(t) &= \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \\
             &= \frac{1}{2}m \left[A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)\right]^2 + \frac{1}{2}k \left[A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)\right]^2 \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)^2 + \cos^2(\omega_0 t + \phi)\right] \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\sin^2(\omega_0 t + \phi) + \cos^2(\omega_0 t + \phi) + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi) + \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)\right] \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[1 + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi)\right] \quad (\text{neglecting the small term } \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)) \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \quad (\text{since } \frac{\gamma^2}{4\omega_0^2} \text{ is very small}) \\
             &= E_0 e^{-\gamma t} \quad \text{where } E_0 = \frac{1}{2}m A_0^2 \omega_0^2 \text{ is the initial energy at } t=0
    \end{align*}
    We can also define the time constant \( \tau \) as the time it takes for the energy to decrease to \( \frac{1}{e} \) of its initial value:
    \begin{equation}
        \tau = \frac{1}{\gamma}
    \end{equation}
    So we have, for very light damping:
    \begin{equation}
        E(t) = E_0 e^{-\frac{t}{\tau}}
    \end{equation}
\end{definition}

@@

\begin{definition}[Quality Factor (Q-Factor)]
    The quality factor \( Q \) is a dimensionless parameter that characterizes the damping of a harmonic oscillator. It is defined as:
    \begin{subequations}
    \begin{equation}
        Q = \frac{\omega}{\gamma} = \omega \tau
    \end{equation}
    And for very light damping, we can approximate \( \omega \approx \omega_0 \), leading to:
    \begin{equation}
        Q \approx \frac{\omega_0}{\gamma} = \omega_0 \tau
    \end{equation}
    \end{subequations}
    This allows us to rewrite the equation of a damped harmonic oscillator as:
    \begin{equation}
        \ddot{x} + \frac{\omega_0}{Q} \dot{x} + \omega_0^2 x = 0
    \end{equation}
    and:
    \begin{equation}
        \omega = \omega_0 \sqrt{1 - \frac{1}{4Q^2}}
    \end{equation}
\end{definition}
\chapter{Modern Physics}
\end{document}

>>>>

\subsection{Energy and Quality Factor}
\begin{definition}[Energy of a Very Light Damping]
    Consider a very lightly damped harmonic oscillator, where \( \gamma \ll \omega_0 \). In this case, the angular frequency of oscillation \( \omega_r \) can be approximated as:
    $$
    \omega_r \approx \omega_0 \left(1 - \frac{\gamma^2}{8\omega_0^2}\right) \approx \omega_0
    $$
    So the motion of the lightly damped oscillator can be approximated as:
    $$
    x(t) \approx A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)
    $$
    Then, we can calculate the velocity of the oscillator:
    \begin{align*}
        \dot{x}(t) &= -\frac{\gamma}{2} A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi) - A_0 \omega_0 e^{-\frac{\gamma}{2}t} \sin(\omega_0 t + \phi) \\
        &= A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)
    \end{align*}
    The total mechanical energy \( E(t) \) of the lightly damped oscillator is given by:
    \begin{align*}
        E(t) &= \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \\
             &= \frac{1}{2}m \left[A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)\right]^2 + \frac{1}{2}k \left[A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)\right]^2 \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)^2 + \cos^2(\omega_0 t + \phi)\right] \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\sin^2(\omega_0 t + \phi) + \cos^2(\omega_0 t + \phi) + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi) + \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)\right] \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[1 + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi)\right] \quad (\text{neglecting the small term } \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)) \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \quad (\text{since } \frac{\gamma^2}{4\omega_0^2} \text{ is very small}) \\
             &= E_0 e^{-\gamma t} \quad \text{where } E_0 = \frac{1}{2}m A_0^2 \omega_0^2 \text{ is the initial energy at } t=0
    \end{align*}
    We can also define the time constant \( \tau \) as the time it takes for the energy to decrease to \( \frac{1}{e} \) of its initial value:
    \begin{equation}
        \tau = \frac{1}{\gamma}
    \end{equation}
    So we have, for very light damping:
    \begin{equation}
        E(t) = E_0 e^{-\gamma t} = E_0 e^{-\frac{t}{\tau}}
    \end{equation}
\end{definition}

@@

\begin{definition}[Quality Factor (Q-Factor)]
    The quality factor \( Q \) is a dimensionless parameter that characterizes the damping of a harmonic oscillator. It is defined as:
    \begin{subequations}
    \begin{equation}
        Q = \frac{\omega}{\gamma} = \omega \tau
    \end{equation}
    And for very light damping, we can approximate \( \omega \approx \omega_0 \), leading to:
    \begin{equation}
        Q \approx \frac{\omega_0}{\gamma} = \omega_0 \tau
    \end{equation}
    \end{subequations}
    This allows us to rewrite the equation of a damped harmonic oscillator as:
    \begin{equation}
        \ddot{x} + \frac{\omega_0}{Q} \dot{x} + \omega_0^2 x = 0
    \end{equation}
    and:
    \begin{equation}
        \omega = \omega_0 \sqrt{1 - \frac{1}{4Q^2}}
    \end{equation}

We can also consider the ratio between the energy at one time and the energy one period later:
\begin{align*}
    \frac{E(t + T)}{E(t)} &= \frac{E_0 e^{-\gamma (t + T)}}{E_0 e^{-\gamma t}} \\
    &= e^{-\gamma T} \approx 1 - \gamma T \\
    \frac{E(t + T) - E(t)}{E(t)} & \approx -\gamma T = -\frac{2\pi}{\omega_0} \gamma = -\frac{2\pi}{Q}
\end{align*}
\end{definition}

\begin{example}
    \textit{What is the number of radians through which the damped system oscillates as its energy decreases to $1/e$ of its initial value?}

    We have
    $$
    \frac{E}{E_0} = e^{-\gamma t} = \frac{1}{e} \implies \gamma t = 1 \implies t = \frac{1}{\gamma} = \tau
    $$
    So the number of radians is
    $$
    \theta = \omega \tau = \frac{\omega}{\gamma} = Q
    $$

\end{example}

\subsection{Undamped Forced Oscillations}
\begin{definition}[Undamped Forced Oscillations]
    Consider a driver porves on the left side of the sping, we have the following differential equation:
    \begin{equation}
        m\ddot{x} + kx = \eta
    \end{equation}
    where \( \eta = F_0 \cos(\omega t) \) is the driving force with amplitude \( F_0 \) and angular frequency \( \omega \). The general solution to this non-homogeneous differential equation given by:
    \begin{equation}
        x(t) = A(\omega)\cos(\omega t - \delta) 
    \end{equation}
    to derive $A(\omega)$ and $\delta$, we can use the equation of motion:
    \begin{align*}
        A(\omega)(-\omega^2 + \omega_0^2) \cos(\delta) = \omega_0^2 a
    \end{align*}
    where \( a = \frac{F_0}{k} \) is the static displacement of the mass when the driving force is constant. We also haev:
    \begin{align*}
        A(\omega)(-\omega^2 + \omega_0^2) \sin(\delta) = 0
    \end{align*}
    Now, consider the case when $\delta = 0$, we have:
    \begin{equation}
        A(\omega) = \frac{\omega_0^2 a}{\omega_0^2 - \omega^2}
    \end{equation}
    and the case where \( \delta = \pi \), we have:
    \begin{equation}
        A(\omega) = -\frac{\omega_0^2 a}{\omega_0^2 - \omega^2}
    \end{equation}
    Observe, when $\omega \approx \omega_0$, the amplitude \( A(\omega) \) becomes very large, indicating resonance. At resonance, the system oscillates with maximum amplitude, which can lead to significant energy transfer from the driving force to the oscillator.
\end{definition}

\subsection{Damped Forced Oscillations}
\begin{definition}[Damped Forced Oscillations]
    Consider a damped harmonic oscillator subjected to an external driving force. It is described by the following differential equation:
    \begin{equation}
        m\ddot{x} + b\dot{x} + kx = F_0 \cos(\omega t)
    \end{equation}
    where \( F_0 \) is the amplitude of the driving force, \( \omega \) is the angular frequency of the driving force, \( b \) is the damping coefficient, \( m \) is the mass, and \( k \) is the spring constant. The general solution to this non-homogeneous differential equation is given by:
    \begin{equation}
        x(t) = A(\omega)\cos(\omega t - \delta)
    \end{equation}
    and the amplitude \( A(\omega) \) is given by:
    \begin{equation}
        A(\omega) = \frac{\omega_0^2 a}{\sqrt{\omega^2 \gamma^2 + (\omega^2 - \omega_0^2)^2}}
    \end{equation}
    We can derive the following 4 cases:
    \begin{enumerate}
        \item \textbf{No Damping} (\( \gamma = 0 \)): In this case, the amplitude \( A(\omega) \) simplifies to the undamped case we discussed earlier.
        \item \textbf{Low-Frequency Limit} (\( \omega \ll \omega_0 \)): In this limit, the amplitude \( A(\omega) \) approaches the static displacement \( a = \frac{F_0}{k} \). This means that at very low frequencies, the system behaves like a static spring, and the mass is displaced by an amount proportional to the applied force.
        \item \textbf{High-Frequency Limit} (\( \omega \gg \omega_0 \)): In this limit, the amplitude \( A(\omega) \) decreases with increasing frequency, following the relation \( A(\omega) \approx \frac{\omega_0^2 a}{\omega^2} \). This indicates that at very high frequencies, the mass cannot respond quickly enough to the rapidly oscillating driving force, resulting in a smaller amplitude of oscillation.
        \item \textbf{Resonance} (\( \omega \approx \omega_0 \)): At resonance, the amplitude \( A(\omega) \) reaches its maximum value but it does not become infinite due to the presence of damping: 
        \begin{equation}
            A_{\text{max}} = \frac{\omega_0^2 a}{\omega_0 \gamma} = \frac{\omega_0 a}{\gamma} = Q a
        \end{equation}
    \end{enumerate}
    We also have the phase shift \( \delta \) given by:
    \begin{equation}
        \tan(\delta) = \frac{\omega \gamma}{\omega_0^2 - \omega^2}
    \end{equation}
    The phase shift \( \delta \) indicates how much the oscillation of the mass lags behind the driving force. The behavior of \( \delta \) can be summarized as follows:
    \begin{itemize}
        \item At low frequencies (\( \omega \ll \omega_0 \)), \( \delta \) approaches 0, meaning the mass oscillates in phase with the driving force.
        \item At resonance (\( \omega = \omega_0 \)), \( \delta = \frac{\pi}{2} \), indicating that the mass oscillates a quarter cycle behind the driving force.
        \item At high frequencies (\( \omega \gg \omega_0 \)), \( \delta \) approaches \( \pi \), meaning the mass oscillates out of phase with the driving force.
    \end{itemize}
\end{definition}

\chapter{Modern Physics}
\end{document}
